---
title: "Linear Algebra"
author: "Andrew P Blake"
date: "October 2020"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

## Modelling in economics **is** linear algebra

That's a bit extreme, but you mostly need to do linear algebra to program up many of the estimators we need, or to solve a rational expectations models. 

## Beginning to program

A few non-linear algebra things we will need are summarized in the following table. 

|                 | R        | Example         | Notes               |
|-----------------|--------------|-----------|---------------------|
|Assign a value   |  `<-`  | `a <- 4`  | Also legal is `a = 4`. But I hate it. | 
|Create a list of values | `c(.)` | `v <- c(1, -2, 22)`  | Defining 'on the fly' |
|Sequence   | `seq(i, k, l)` | $5$, $7$, ... ,$21$    | Create a sequence |
|           | `i:k` | $i$, $i\pm 1$, ... ,$k$ | Short cut for unit in/de-crements |
|Loop commands | `for (var in seq) expr` | `for (i in 5:1) print(i)` | Loops. We need loops.|
|Draw a random number | `rnorm(k,a,b)` | `rnorm(60, 0, 5)` | Example draws 60 values ~ $N(0,5)$ |
|Create a matrix | `matrix(v,i,j)` | `matrix(5, 2, 2)` | Create a $2\times 2$ matrix of 5s |

Table: Useful things

### Functions

Everything in R is a function (although it doesn't look like it). Defining a function is simple:
```{r echo=TRUE, eval=FALSE}
name_of_function <- function(function_arguments){
  # Body of function where stuff is done  
}
```

Here's one that actually does something:
```{r echo=TRUE}
addaddadd <- function(x,y){
  z <- 3*(x+y)
  return(z)
}
```
and if we run it:
```{r echo=TRUE}
addaddadd(4,6)
```

## Linear algebra

Assume the following: $A$ and $B$ are real matrices of dimension $n\times n$, $b$ and $c$ are real $n-$vectors, $X$ is a real $T\times k$ matrix, and $S$ is a symmetric real matrix.

|                 | Maths        | R         | Notes               |
|-----------------|--------------|-----------|---------------------|
|Hadamard product | $A\bigodot B$ | `A * B`   | Element-by-element, $A$, $B$ same size |
|Matrix/vector product | $A\times B$, $A \times b$   | `A %*% B`, `A %*% b`   | Normal product rule |
|Inner product    | $X'X$        | `t(X) %*% X` | Also uses transpose operator, `t()` |
|                 |              | `crossprod(X)` | More efficient, but less mathy|
|                 | $A'B$        | `t(A) %*% B` | |
|                 |              | `crossprod(A,B)` |  |
|Outer product    | $A\times B'$ | `tcrossprod(A,B)` | |
|Inverse          | $A^{-1}$     | `solve(A)` | Matrix inverse is a special case of... |
|Solve for $d$    | $Ad = b \Rightarrow d = A^{-1}b$     | `d <- solve(A, b)` | ...linear solution! |
|Cholesky decomp  | $S = R'R$    | `R <- chol(S)` | $S$ is a symmetric, positive definite matrix |
|Cholesky inverse | $S^{-1}$     | `chol2inv(chol(S))` | Fast! |
|Determinant | $\vert A \vert$ | `det(A)` | |
|Diagonal         |              |             |   | 
| $\quad$ of a matrix |   | `diag(A)`   | Retrieve the elements $a_{ii}$, $i=1,..,n$ |
| $\quad$ in a matrix |   | `A <- diag(b)` | Set the diagonal of $A$ to $b$, zero elsewhere | 
| $\quad$ Identity matrix | $I_n$ | `diag(n)` |  | 
|Eigenvalues/vectors     |       | `E <- eigen(A)` | Returns a _list_: `E$values`, `E$vectors` |

Table: Maths commands essential to linear algebra

## Starting to do linear algebra

### Problem

Consider the following simultaneous system of equations:
$$
\begin{align}
x_1 + 2x_2 &= 6 \\
x_1 - 3x_2 +2 x_3 &= 0 \\
-2 x_1 + 3 x_3 &= 2
\end{align}
$$
Find the values of $x$ that solve this using R.

**Hint** -- write the problem in matrix form 
$$
 Ax = b
$$
where
$$
A = \left [ \matrix{1 &2 & 0 \cr
             1 & -3 & 2\cr
             -2 & 0 & 3} \right ], \qquad 
  b = \left[ \matrix{6 \\ 0 \\ 2} \right ]
$$
and then use `solve`.

### Solution

R code to create these matrices is:
```{r}
A <- matrix(c(1,1,-2,2,-3,0,0,2,3),3,3) # Matrices are populated by column by default
b <- matrix(c(6,0,2),3,1)
```
The solution is:
```{r}
x <- solve(A,b)
```
where `x` is:
```{r echo = FALSE}
x
```

## Eigenvalue decomposition

Any square matrix can be decomposed into a non-singular matrix $V$ of eigenvectors and a diagonal matrix of eigenvalues $\Lambda$ such that:
$$
A V = V \Lambda \Rightarrow A = V\Lambda V^{-1}
$$

Call `eigen` to decompose our previously defined matrix $A$
```{r}
e <- eigen(A)
L <- e$values    # Recall a list is returned so assign vectors/values
V <- e$vectors
```
Note $A$ is **not** symmetric so it may have complex roots, which is does
```{r}
L
```
If we calculate $A = V\Lambda V^{-1}$ we get 
```{r}
s <- V %*% diag(L) %*% solve(V)
s
```
and when we drop the zero imaginary parts 
```{r}
Re(s)
```
and then round we get
```{r}
round(Re(s), digits=12)
```

## Precision

Take a real matrix $A_{mn}$ with $n \le m$ and pre-multiply by its own transpose, i.e.\ $AA = A'A$. $AA$ is symmetric, positive semi-definite. If $rank(A) = n$, then $rank(AA) = n$ and positive definite, and its inverse exists.

```{r}
A  <- matrix(c(1, .2, 0, 1), 2, 2)
AA <- t(A) %*% A
AA
```
Let's invert $AA$ three different ways.
```{r}
i1 <- solve(AA) 
i2 <- chol2inv(chol(AA))
i3 <- qr.solve(AA) 

i1 %*% AA
i2 %*% AA
i3 %*% AA
```

```{r}
id <- i3 %*% AA
T1 <- knitr::kable(as.data.frame(id), caption = "An identity matrix", format="html")
kableExtra::kable_styling(T1, full_width = FALSE, position = "center")
```

## Programming the regression problem

Let's look at the familiar regression problem for some generated data.
$$
  y = XB + \epsilon
$$
where $\epsilon \sim N(0,.2)$, $X$ is a $(k+1)\times n$ matrix of regressors including a constant and $B$ a $k+1$ vector of coefficients. Let's generate some random data of an arbitrary sized problem:
```{r manip}
X <- matrix(rnorm(180, 2, 1), 60, 3)
head(A, 6) # Print first six rows
X <- cbind(1, X) # Add a constant
tail(X,6) # Print last six rows
```

Now create a dependent variable that is a linear combination of these variables plus some noise. Create the linear relationship first so we know what it is:
```{r B}
B <- matrix(c(0.5,1,-1,.2), 4, 1)
```
and then the dependent variable:
```{r dep}
y <- X %*% B + 0.2*rnorm(60)
```

We could now do a regression - i.e. calculate
$$
  \hat B = (X'X)^{-1}X'y
$$
which can be written:
```{r reg}
Bhat <- solve(t(X)%*%X)%*%t(X)%*%y
```
which gives
```{r reg2, echo=FALSE}
Bhat
```
But I wouldn't do it like this (we'll see why in a minute). A better way would be 
```{r better}
Bhat2 <- chol2inv(chol(crossprod(X)))%*%crossprod(X,y)
```
or even
```{r betteryet}
Bhat3 <- qr.solve(X,y)
```
which both evaluate to the same $\hat B$ values.

## Testing timings

Why does it matter how you do things? It should be obvious that it might, but it turns out some fairly trivial things can make a lot of difference. We set some parameters so we can create a bigger problem.
```{r message=FALSE, warning=FALSE}
n <- 400
k <- 12
```

We will use seven different methods to calculate an estimate of $B$. These are two variations on the three calculations below (where the brackets matter!):
$$
\hat B_1 = ((X'X)^{-1}) X' y 
$$

$$
\hat B_2 = ((X'X)^{-1}) (X' y) 
$$
$$
\hat B_3 = ((X'X)^{-1} (X' y))
$$
where we do it either 'by hand' or using `crossprob`, plus using `qr.solve`.


```{r}
library(tictoc)

reps <- 10000
t    <- list()

set.seed(123)
tic()
for (i in 1:reps) { 
  B <- matrix(runif(k+1), k+1, 1)
  X <- cbind(1L, matrix(rnorm(n*k, 2, 1), n, k))
  y <- X %*% B + 0.2 * rnorm(n) 
  Bhat <- solve(t(X) %*% X) %*% t(X) %*% y 
  }
t[[1]] <- toc(quiet = TRUE)

set.seed(123)
tic()
for (i in 1:reps) { 
  B <- matrix(runif(k+1), k+1, 1)
  X <- cbind(1L, matrix(rnorm(n*k, 2, 1), n, k))
  y <- X %*% B + 0.2 * rnorm(n) 
  Bhata <- solve(t(X)%*%X) %*% (t(X)%*%y) 
}
t[[2]] <- toc(quiet = TRUE)

set.seed(123)
tic()
for (i in 1:reps) { 
  B <- matrix(runif(k+1), k+1, 1)
  X <- cbind(1L, matrix(rnorm(n*k, 2, 1), n, k))
  y <- X %*% B + 0.2 * rnorm(n) 
  Bhatb <- solve((t(X) %*% X), (t(X) %*% y)) 
  }
t[[3]] <- toc(quiet = TRUE)

set.seed(123)
tic()
for (i in 1:reps) { 
  B <- matrix(runif(k+1), k+1, 1)
  X <- cbind(1L, matrix(rnorm(n*k, 2, 1), n, k))
  y <- X %*% B + 0.2 * rnorm(n) 
  Bhat2 <- solve(crossprod(X)) %*% crossprod(X,y) 
  }
t[[4]] <- toc(quiet = TRUE)

set.seed(123)
tic()
for (i in 1:reps) { 
  B <- matrix(runif(k+1), k+1, 1)
  X <- cbind(1L, matrix(rnorm(n*k, 2, 1), n, k))
  y <- X %*% B + 0.2 * rnorm(n) 
  Bhat2a <- solve(crossprod(X), crossprod(X,y)) 
  }
t[[5]] <- toc(quiet = TRUE)

set.seed(123)
tic()
for (i in 1:reps) { 
  B <- matrix(runif(k+1), k+1, 1)
  X <- cbind(1L, matrix(rnorm(n*k, 2, 1), n, k))
  y <- X %*% B + 0.2 * rnorm(n)
  Bhat2b <- chol2inv(chol(crossprod(X))) %*% crossprod(X,y) 
  }
t[[6]] <- toc(quiet = TRUE)

set.seed(123)
tic()
for (i in 1:reps) { 
  B <- matrix(runif(k+1), k+1, 1)
  X <- cbind(1L, matrix(rnorm(n*k, 2, 1), n, k))
  y <- X %*% B + 0.2 * rnorm(n) 
  Bhat3 <- qr.solve(X,y) 
  }
t[[7]] <- toc(quiet = TRUE)
```

How do we display the timings we saved in `t`? We could do a simple (but dull) table, or something a but nicer.

```{r warning=FALSE, message=FALSE, fig.align="center"}
library(tidyverse)

FF <- c("solve(t(X)%*%X) %*% t(X) %*% y",
        "solve(t(X)%*%X) %*% (t(X)%*%y)",
        "solve(t(X)%*%X, (t(X)%*%y))",
        "solve(crossprod(X)) %*% crossprod(X,y)",
        "solve(crossprod(X), crossprod(X,y))",
        "chol2inv(chol(crossprod(X))) %*% crossprod(X,y)",
        "qr.solve(X,y)")

v <- tibble(Method = 1:7, 
            Times = unlist(t)[seq(2,14,2)] - unlist(t)[seq(1,13,2)], 
            Formula = FF) %>% 
  mutate(Col = case_when(
    Times == min(Times) ~ "red",
    TRUE ~ "blue"
  ))

ggplot(v) + 
  geom_col(aes(x=Method, y=Times, fill=as.factor(Method)), alpha = .6) + 
  geom_text(aes(x=Method, y=.1, label=Formula, color=Col), size=5.5, hjust=0) +
  theme_minimal() + 
  scale_color_identity() + 
  theme(legend.position = "none") + 
  theme(axis.text.y = element_blank()) + 
  coord_flip() + 
  labs(y=paste("Seconds taken to do",reps,"replications"), x="", 
       title="Timings of different numerical regression methods")
```


